Improving the prospects for multimessenger astronomy with \earlywarning\ detection of compact binary coalescence.

### 2011-03-31 method.tex
substantial rework, moved comp. cost portion of appendix into main body
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@711 ccdd05d5-56c1-461a-bf98-85f6876523b4

\section{Method}
\label{SECII}\label{sec:method}
\subsection{The standard approach: matched filtering}

Searches for gravitational waves from compact binary coalescences typically
employ matched filter banks \cite{findchirppaper}.  Potential inspiral signals
are continuously parameterized by time, amplitude, phase, and a set of
intrinsic source parameters $\theta$, which in this paper we shall take to
consist of the two component masses of a binary, $\theta = (m_1, m_2)$.  Let
$h_+(\theta)$ and $h_\times(\theta)$ be, respectively, the `$+$' and `$\times$'
polarization gravitational wave signals that would arise from a fiducial
face-on binary at some distance.  Because, for inspiral signals, $h_+$ and
$h_\times$ are nearly in quadrature, they are generally combined into a single
complex-valued template $h = h_+ + i h_\times$.

The detection procedure for just one set of intrinsic source parameters
$\theta$ starts by {\em whitening} the measurement data stream $x$.  This
involves finding a linear filter that renders the detector's noise \textsc{iid}
and Gaussian.  This filter is applied to the measurement, yielding the whitened
data stream $x^\mathrm W$.  The same linear filter is applied to the template
$h(\theta)$, yielding the whitened template $h^\mathrm W (\theta)$.  The
matched filter is the normalized cross-correlation of $h^\mathrm W$ and
$x^\mathrm W$, \editorial{After introducing the $\star$ notation for cross
correlation, must we define it for continuous and discrete time?}
%
$$
%
\rho(\theta) = \frac{h^\mathrm W (\theta) \star x^\mathrm W}{|h^\mathrm W
(\theta)|}.
%
$$
%
This is called the signal to noise ratio, or \textsc{snr}.  The detection
statistic is the modulus of this, $|\rho(\theta)|$, which has a $\chi^2$
distribution with 2 degrees of freedom in the absence of signal.

To construct a filter bank, matched filters are realized for discrete signal
parameters $\theta_1,\, \theta_2,\, \cdots$, $\theta_N$, such that any possible
signal will have a maximum cross-correlation of at least 0.97 with at least one
template.  \editorial{Citation needed for template placement procedure?}  Such
a template bank is said to have a 97\% {\em minimum match}.  This technique is
designed so that an inspiral signal can be detected without any prior knowledge
of its intrinsic parameters: at most 3\% of the \textsc{snr} is lost by a
signal's parameters not exactly coinciding with a template's.  A trigger is
reported for the template parameters $\theta_i$ and time $t$ for which $|\rho|$
is a maximum over some moving interval in $\theta$ and $t$.

\subsubsection{Latency and overhead}

The matched filter bank can be implemented with finite impulse response
(\textsc{fir}) filters.  \textsc{fir} filters are perfectly suited for realtime
detection, because they do not introduce any latency at all.  However,
\textsc{fir} filters are very expensive: for $N$ templates of $M$ samples each,
a \textsc{fir} matched filter bank costs $\mathcal O(M N)$ operations per
sample.

Much more commonly, the matched filter bank is implemented using \textsc{fft}
convolutions, costing only $\mathcal O(M \lg N)$ \editorial{$\lg$ is $\log_2$.
Explain nomenclature? Switch to $\log_2$?} per sample but having a latency of
at least $M$ samples.  For example, for a bank of $1\,\mathrm{ks}$ templates
sampled at $4096\,\mathrm{Hz}$, the \textsc{fft} implementation requires about
about $2.2 \times 10^4$ times fewer floating point operations per sample than
the \textsc{fir} implementation.  However, the \textsc{fft} implementation has
a latency of at least $1\,\mathrm{ks}$.

This presents a dilemma: it seems that low latency detection using the
\textsc{fir} implementation is prohibitively expensive, whereas computationally
cheap detection with the \textsc{fft} method comes with minutes to hours of
latency.

In the rest of the section, we will describe a detection strategy that makes
use of some very general properties of inspiral template banks in order to
evaluate a matched filter bank with far lower computational cost than the
conventional \textsc{fir} method and far less latency than the conventional
\textsc{fft} method.

@@

Our first innovation is to split each template into disjoint intervals, or
\emph{time slices}.  A matched filter is constructed for each time slice, whose
outputs form an ensemble of partial \textsc{snr} streams.  By linearity, these
partial \textsc{snr} streams can be suitably time delayed and summed to
reproduce exactly the \textsc{snr} with respect to the original template.  By
exploiting some general properties of inspiral waveforms, we shall see that
time slices permit a reduction in latency relative to the \textsc{fft} matched
filter, but with comparable computational overhead.

\editorial{\textsc{mbta} should be cited in the introduction.}  A similar idea
has been demonstrated by the Virgo Collaboration's \textsc{mbta}
pipeline~\cite{Marion2004,beauville2006,beauville2008,Buskulic2010}, which
operated in a low-latency mode during \textsc{ligo}'s sixth science run and
Virgo's second science run in 2010.

Each partial matched filter can be realized with either the \textsc{fir} or the
\textsc{fft} method.  If all of the time slices make use of the \textsc{fft}
method, the resulting filter will have lower latency than a conventional
\textsc{fft} matched filter, and also less computational overhead than an
\textsc{fir} matched filter.  A further benefit is that all of the time slices
may be processed in parallel for additional speedup.

If the template is known to be quasi-bandlimited, then the partial matched
filter for each time slice may be processed at a reduced sample rate.  Compact
binary inspiral waveforms fit this description superbly: they are chirps whose
frequency and amplitude rise according to power laws of time.  They are not
only quasi-bandlimited, but quasi-monochromatic.  By exploiting this property,
we can implement a matched filter for an inspiral waveform using a multirate
filter bank that requires far fewer floating point operations.

For concreteness and simplicity, let us consider an inspiral waveform in the
quadrupole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{-t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \mu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992, blanchet2002,
hanna2009}. \editorial{Way too many citations here.} Usually the template is
truncated at some prescribed time $t = t_0$, often chosen to correspond to the
innermost stable circular orbit at which $f = f_\mathrm{ISCO} \approx 4400 \,
M_\odot / M$.  An inspiral signal will enter the \textsc{ligo} band at a low
frequency cutoff, $f = f_\mathrm{low}$ corresponding to a time
$t_\mathrm{low}$.  This template has a duration of $t_\mathrm{0} -
t_\mathrm{low}$ and is critically sampled at a rate of $2 f_\mathrm{isco}$.

The time slices for this template consist of the $K$ intervals $(t_K, t_{K-1}],
\dots, (t_2, t_1], (t_1, t_0]$ sampled at frequencies $f_\mathrm{K-1}, \dots,
f_1, f_0$ where $f_0 = 2 f_\mathrm{ISCO}$, $t_0 = t_\mathrm{ISCO}$, $f_{K-1}
\geqslant 2 f_\mathrm{low}$, and $t_K \leqslant t_\mathrm{low}$.

If all of these time slices are implemented using the \textsc{fft}, then the
latency of this filter bank is $\max_j {\left[-\left( 2(t_j - t_{j-1}) -
t_{j-1} \right)\right]} = \max_j \left( 3 t_{j-1} - 2 t_j \right)$.

\begin{table}[h!]
\caption{Example of critically sampled, power-of-2 time slices for a $1.4 - 1.4
\, M_{\odot}$ template extending from $f_\mathrm{low} = 10 \, \mathrm{Hz}$ to
$f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$ with a time frequency structure given by
($\ref{eq:fgw})$.}
\label{table:time_slices}
\begin{minipage}[c]{0.5\textwidth}
\centering
\includegraphics[trim = 3cm 0cm 1.5cm 0cm, scale=0.4]{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\end{table}

An example time slice design satisfying these constraints for a $1.4 - 1.4 \,
M_{\odot}$ is shown in table~\ref{table:time_slices}.  For this example, the
latency for this time slice design is just $\input{time_slice_latency.tex}$
even if all of the time slices are implemented with the \textsc{fft} method.
This set of time slices will require $\input{time_slice_ops_firslice.tex}$
operations per sample per template, compared with
$\input{time_slice_ops_conv.tex}$ for pure \textsc{fft} cross-correlation
without time slices, or $\input{time_slice_ops_td.tex}$ for the \textsc{fir}
filter method without time slices.

@@

Our second innovation exploits the fact that the templates in inspiral template
banks are, by design, highly correlated.  This is due to the fact that points
are chosen in parameter space to be templates such that any point from the
region parameter space we cover has a match to the nearest template greater
than our chosen minimum match. It is possible to greatly reduce the number of
matched filters required to achieve a particular minimum match by designing an
appropriate set of orthonormal {\em basis templates}.  A purely numerical
technique based on the application of the singular value decomposition
(\textsc{svd}) to inspiral waveforms is demonstrated by the authors
in~\cite{Cannon:2010p10398}.

One may regard a bank of $N/2$ discretely-sampled, complex-valued templates of
length $M$ samples, $h^\mathrm W(\theta_i; t_j) = [\mathbf H]_{ij} = H_{ij}$,
as a real matrix of shape $N \times M$, where we have unpacked each
complex-valued template into two real-valued templates.  The singular value
decomposition is an exact factorization that exists for any matrix such that
%
\begin{equation} \label{eq:svd-exact}
%
H_{ij} = [\mathbf {V \Sigma U}]_{ij} = \sum_{k=1}^{N} v_{ik} \sigma_k u_{kj},
%
\end{equation}
%
where $\mathbf V$ and $\mathbf U$ are both unitary matrices and $\mathbf
\Sigma$ is a diagonal matrix.  For our purposes, we associate the rows of
$\mathbf U$ with a minimal set of basis templates, which become the kernels of
basis filters.  These filters give rise to the orthogonal \textsc{snr}s,
$\rho_k^\perp = u_k \star x^\mathsf{W}$.  The rows of $\mathbf{V \Sigma}$
become reconstruction coefficients that map linear combinations of the
orthogonal \textsc{snr}s from the basis filters back onto \textsc{snr}s for the
original templates of interest: $\rho_i = \sum_k v_{ik} \sigma_k \rho_k^\perp$.

In many applications of the \textsc{svd}, including ours, the matrix can be
well approximated by truncating the summation in equation~(\ref{eq:svd-exact})
at $N' \ll N$:
%
\begin{equation}
%
H_{ij}^\prime = \sum_{k=1}^{N'} v_{ik} \sigma_k u_{kj}
%
\end{equation}
%
The cumulative sum of squares of the {\em singular values}, $\sigma_k$,
measures the Frobenius norm of the approximation, such that
%
\begin{equation}
%
\frac{\| \mathbf H^\prime \|}{\| \mathbf H \|} = \left(\sum_{k=1}^{N'}
|\sigma_k|^2\right)^{1/2} \left(\sum_{k=1}^{N} |\sigma_k|^2\right)^{-1/2}.
%
\end{equation}
%
This is also called the \textsc{svd} tolerance.  In our application, it relates
to how much \textsc{snr} is lost by discarding $N - N'$ of the basis filters
with the lowest singular values.

This result differs from other work that models gravitational-wave chirp
signals in approximate ways~\cite{Chassande-Mottin2006, Candes2008,
BuonannoChenVallisneri:2003a} by starting with an exact representation of the
desired template family and producing a rigorous approximation with a tunable
accuracy.

\subsection{Composite detection statistic and hierarchical detection}

\editorial{We should cite \cite{Scharf:1994p12497} here.} Although the
\textsc{svd} allows us to reduce the number of matched filters, this comes at
the price of having to perform a matrix multiplication by the reconstruction
matrix $\mathbf {V \Sigma}$ at every sample.  In some circumstances, this
matrix multiplication may be more expensive than applying the orthogonal
matched filters.

Further speedup may be gained in a hierarchical detection scheme.  In general,
the orthogonal \textsc{snr}s alone provide some indication of whether any
template in the original template bank is likely to have a large \textsc{snr}.
Consider some composite detection statistic that is a scalar function of the
orthogonal \textsc{snr}s, $\Gamma(\rho_1^\perp, \rho_2^\perp, \cdots,
\rho_L^\perp)$.  Suppose that we can infer the distribution of $\Gamma$ for a
data stream with the signal present or with the signal absent.  Then we may use
a threshold crossing of $\Gamma$ to trigger the conditional application of the
expensive reconstruction matrix only during the times when a signal is likely
to be present.

Following the ideas in~\cite{Anderson:2000yy}, we employ is a weighted sum of
squares, $\Gamma = \sum_{k=1}^{N'} w_k (\rho_k^\perp) ^ 2$, with the particular
weights
%
\begin{equation}
%
w_k = \frac{\sigma_k^2}{\sigma_k^2 + N / A^2},
%
\end{equation}
%
where $A^2$ is a desired \textsc{snr} scale that is set by the analyst.  As the
authors show in~\cite{svd-compdetstat}, this choice of weights has a better
receiver operating characteristic for a signal of \textsc{snr}~=~$A$ than some
other obvious choices, $w_k = \sigma_k^2$ or $w_k = 1$.

\subsection{Comparison of computational costs and latency}

Using all of these innovations, we are able to greatly reduce the compuational
cost of searching for these signals. An example of this can be seen in
table~\ref{table:comp_cost} where we compute the computational cost of
filtering 256 templates from a $97\%$ minimal match template bank whitened with
an (FIXME: is this correct?) Advanced \textsc{ligo} \textsc{asd}. It is
interesting to note that by themselves, both the \textsc{svd} and time-slice
innovations \emph{increase} the overall computational cost when combined with
\textsc{fft} filtering. It is only when combined with conditional
reconstruction do these methods yield computational savings. Details of the
computational cost of different operations can be found in \ref{appendix:cost}.

\editorial{It would be good to illustrate the layout of this particular
template bank: masses spanned, time slice layout \dots}
\begin{table}[!h]
\caption{Operation counts per sample for six different detection methods.  The
operation counts for \textsc{lloid} assume a reconstruction duty cycle of 5\%.
Note that the \textsc{fft} method with \textsc{lloid} is almost 10 times faster
than the conventional \textsc{fft} method, despite having substantially lower
latency.}
\label{table:comp_cost}
\begin{center}
\input{flop_budget}
\end{center}
\end{table}

>>>>

\section{Early warning searches for compact binary coalescence}
\label{SECII}\label{sec:method}

In this section we describe a decomposition of the compact binary parameter
space that reduces low latency filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
search with $\sim$minutes latency in LIGO's S6 and Virgo's VSR2 science runs.
We combine this with the orthogonal decomposition described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

Inspiral signals are parameterized by a set of intrinsic parameters
$\bar{\theta}$ that determine the amplitude and phase evolution of the observed
binary signals.  Searches for gravitational waves from compact binary
coalescence typically employ matched filter banks, called template banks,
\cite{findchirppaper} that discretely sample the possible intrinsic parameters.
The filter models of the waveforms, known as templates are weighted by the
detector noise amplitude spectral density, $x(t,\bar{\theta})$ and filtered
against the whitened detector data.  A sufficient number of templates are
chosen to assure a minimal loss of SNR~\cite{Owen:1995tm,Owen:1998dk}.  To
construct a template bank, matched filters are chosen with discrete signal
parameters $\theta_1,\, \theta_2,\, \cdots$, $\theta_N$, such that any possible
signal will have a cross-correlation of $\geq0.97$ with at least one template.
Such a template bank is said to have a 97\% {\em minimum match}. For systems
where the effects of spin can be ignored, the intrinsic parameters are the
component masses of the binary, $\bar{\theta} = (m_1, m_2)$.

The procedure for choosing a bank of matched filters for gravitational wave
searches is known~\cite{Owen:1995tm}. Therefore, here we will assume that a
bank of filters has already been chosen based on the targetted parameter space
and we will denote the $i^{\mathrm th}$ filter with parameters $\bar{\theta_i}$
as a function of time $x_i(\tau)$.%
%
\footnote{There are two gravitational wave polarizations, $+$ and $\times$. A
given detector will observe a combination of these polarizations that will
largely be degenerate with an overall unknown constant phase.  This can be
maximized over by filtering for quadrature phases and taking the magnitude of
the result. For simplicity we will ignore that aspect in this work as it is
straightforward to generalize, but not necessary for understanding any of the
points that will be made.}
%
In this work we will discuss transformations to a set of filters
$\{x_i(\tau)\}$.  Some of these transformations are not useful or practical
over the entire parameter space.  For that reason we assume from here onward
that the set of filters $\{x_i(\tau)\}$ refer to a set of near neighbor filters
that can be chosen as a subset of the full parameter space.  Several such
subsets can be chosen until all of the filters are a member of one local set.

Filtering the detector data $h(t)$ involves a convolution of the data with the
filter.  For a unit-normalized filter, and whitened detector data, the result
can be interpreted as the signal-to-noise ratio, $\rho_i(t)$ and is defined as
%
% Filtering equations
%
\begin{eqnarray}
\rho_i(t) &=& \int x_i(\tau) \, h(t-\tau) \, d \tau \label{eq:SNRTD} \\
          &=& \int \tilde{x}_i(f) \, \tilde{h}(f) \, \e^{-2\pi\i ft} \label{eq:SNRFD} df,
\end{eqnarray}
%
%
where the second line is a result of the convolution theorem and
$\tilde{x}_i(f)$ is the Fourier transform of $x_i(\tau)$ as is $\tilde{h}(f)$ the
Fourier transform of $h(t)$.

The evaluation of the integrals in \eqref{eq:SNRTD} and \eqref{eq:SNRFD} are
implemented as sums over sample points for the digitized gravitational wave
detector output.  Discrete Fourier transforms can be computed efficiently
numerically.  For that reason \eqref{eq:SNRFD} is typically far faster,
computationally.  To evaluate \eqref{eq:SNRTD} requires \order{\tmpsamps
\hoftsamps} floating point operations per filter $x_i$, where \tmpsamps\ is the
number of sample points in the filter $x_i$ and \hoftsamps\ is the number of
sample points in the data $h$.  Assuming \numtmps\ filters are required, the
total cost is \order{\numtmps \tmpsamps \hoftsamps} However, \eqref{eq:SNRFD}
requires only \order{\hoftsamps \log \hoftsamps} operations per filter assuming
transform lengths that are longer than the filter (i.e. \hoftsamps\ $>$
\tmpsamps), resulting in a total cost of \order{\numtmps \hoftsamps \log
\hoftsamps}.  In most cases \tmpsamps\ $\gg$ $\log$ \hoftsamps\ and the
computational savings by choosing the frequency domain integral form
\eqref{eq:SNRFD} is clear.  However, to take full advantage of the
computational efficiency of \eqref{eq:SNRFD} requires an acausal knowledge of
the detector data $h(t)$, which implies an inherent latency.  In contrast,
\eqref{eq:SNRTD} can be updated every time a new sample point of detector data
is taken.

In order to minimize latency we propose using the time domain convolution
presented in \eqref{eq:SNRTD}.  However, because the brute force evaluation of
\eqref{eq:SNRTD} is far too costly to be useful, we will consider an
approximation to \eqref{eq:SNRTD} that can reduce substantially the cost of
\realtime\ filtering. This approximation has the form
%
% orthogonal decomposition filtering
%
\begin{equation}
\rho_i(t) \approx \sum_k^{\numslices} \sum_j^{\numsvdtmps} 
	\int_{\tau_{k}}^{\tau_{k+1}} v_{ijk} \sigma_{jk} u_{jk}\,(\tau) \, h(t-\tau) \, d \tau \label{eq:decomp}
\end{equation} 
%
%
where $u_{jk}(\tau)$ is an orthogonal basis set of filters spanning the space
of $\{x_i(\tau)\}$ and $\sigma_{jk} v_{ijk}$ is a tensor relating the filters
$u_{jk}(\tau)$ to the original filter set $\{x_i(\tau)\}$.  We claim that with
a suitable choice of filters $u_{jk}(\tau)$ one can reduce the computational
cost of \eqref{eq:SNRTD} sufficiently to feasibly search for
gravitational-waves from compact binary coalescence in \realtime.  This
requires 1) exploiting the redundancy of the template bank and 2) exploiting
the time-frequency characteristics of the binary waveforms.  We describe our
procedure for producing the decompostion in \eqref{eq:decomp} in the remainder
of this section.

@@

The first step of the orthogonal decomposition described in \eqref{eq:decomp}
is to divide the templates into \emph{time slices}.  This is a time domain
analogue to the frequency domain decomposition described in ~\cite{Marion2004,
Buskulic2010, beauville2006, beauville2008}.  A matched filter is constructed
for each time slice.  The outputs form an ensemble of partial \textsc{snr}
streams.  By linearity, these partial \SNR\ streams can be suitably time
delayed and summed to reproduce the \SNR\ of the full template.  We will show
in the next section that this, combined with the singular value docomposition,
is sufficient to enable a computationally efficient time-domain search and
furthermore is an essential part of an \earlywarning\ detection scheme.

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrupole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{-t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \mu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t_0$, or equivalently frequency \fmax.
This is often chosen to correspond to the \ISCO\ defined previously. An inspiral signal
will enter the detection band at a low frequency, $f = f_\mathrm{low}$
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t_\mathrm{low}, t_\mathrm{0})$ and is said to have  a
duration of $t_\mathrm{0} - t_\mathrm{low}$. It is critically sampled at a
rate of $2 \fmax$.

The monotonic time-frequency relationship of \eqref{eq:fgw} allows us to choose
time slice boundaries that require substantially less bandwidth at early times
in the inspiral.  Our goal is to reduce the filtering cost of a large fraction
of the waveform by computing part of the filter integral at a lower sample
rate.  Specifically we consider here time slice boundaries with the next
highest power-of-two sample rate that critically samples the time sliced
filter.  The time slices for this template consist of the $k$ intervals $(t_k,
t_{k-1}], \dots, (t_2, t_1], (t_1, t_0]$ sampled at frequencies
$f_\mathrm{k-1}, \dots, f_1, f_0$ where $f_0 \geq 2 f_\mathrm{ISCO}$, $t_0 =
t_\mathrm{ISCO}$, $f_{k-1} \geqslant 2 f_\mathrm{low}$, and $t_k \leqslant
t_\mathrm{low}$. An example time slice design satisfying these constraints for
a $1.4 - 1.4 \, M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time frequency structure given by ($\ref{eq:fgw})$. $f_k$ is the sample
rate of the time slice, $(t_{k+1}, t_k]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ are the number of sample points in the
$k^{\mathrm{th}}$ filter.}
\end{table}
%

Rather than applying a unique time slice decomposition to each template
waveform, we find a decomposition that is adequate for the entire set
$\{x_i\}$.  This is facilated by choosing templates in the set $\{x_i\}$ with
similar chirp masses, $\mathcal{M}$.  The time slice decompositions of the
filters $x_i(\tau)$ lead to new filters $y_{ik}(\tau)$ satisfying
%
\begin{equation}
\label{eq:time_slices}
x_i(\tau) \approx \sum_k^{\numslices} y_{ik}(\tau)
\end{equation}
where \numslices\ is the number of time slices required.  Note that we write
the relationship as approximate since the resampling implementation may have
some inherent loss in quality.  We note that by construction these filters are
orthogonal over the index $k$ since they are disjoint in time.  In the next
section we examine how to reduce the number of filters $y_{ik}(\tau)$ via
singular value decomposition to construct a set of filters that is also
orthogonal within a given time slice.

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of orthonormal {\em
basis templates}.  A purely numerical technique based on the application of the
singular value decomposition (\SVD) to inspiral waveforms is demonstrated
in~\cite{Cannon:2010p10398}.  Using the results of ~\cite{Cannon:2010p10398} we
establish that the filters of the previous section can be approximated to high
accuracy by the expansion in the singular value basis
%
\begin{equation}
y_{ik}(\tau) \approx \sum_j^{\numsvdtmps} \sigma_{jk} v_{ijk} u_{j}(\tau)
\label{eq:svddecomp}
\end{equation}
%
where $\sigma_{jk}$ is the $j^{\mathrm{th}}$ singular value for the
$k^{\mathrm{th}}$ time slice, $v_{ijk}$ is an orthogonal matrix for the
$k^{\mathrm{th}}$ time slice  and $u_{jk}$ is a new orthogonal basis filter set
for the $k^{\mathrm{th}}$ time slice.  The authors of \cite{Cannon:2010p10398}
showed that to high reconstruction accuracy far fewer filters are needed than
were in the original template bank. We find that when combined with the time
slice decomposition, the number of \SVD\ filters, \numsvdtmps\ is much smaller
than the original number of filters \numtmps.  We combine \eqref{eq:svddecomp}
with \eqref{eq:time_slices} to arrive at \eqref{eq:decomp}.  In the next
section we compute the expected computational cost scaling of this
decomposition and compare it with the brute force implementation in
\eqref{eq:SNRTD} and higher latency \fft\ methods.

\subsection{Comparison of computational costs}

We now examine the computational cost scaling of the approximate implementation
of \eqref{eq:SNRTD} as \eqref{eq:decomp}.  An actual implementation of this
decomposition in a working analysis pipeline is discussed in the next section
along with measured computational requirements.  For convenience, table
\ref{tab:recap} is a recap of the meaning of various symbols used in this
calculation.
%
% symbols used in FLOPs calculations table
%
\begin{table}
\begin{tabular}{rl}
\bf{Symbol}	& \bf{Definition} \\
\hline
\hoftsamps	& The number of sample points in the data \\
\tmpsamps	& The number of sample points in template $x_i$ \\
\numtmps	& The number of templates in the set $\{x_i\}$ \\
\numslices	& The number of time slices \\
\numsvdtmps	& The number of orthogonal filters in time slice $k$ \\
\slicessamps	& The number of samples in time slice $k$ \\
\fmax		& The maximum frequency of a filter \\
$f_k$		& The sample frequency of the $k^{\mathrm{th}}$ filter \\
\resampsamps	& The number of sample points in the resample filter
\end{tabular}
\caption{\label{tab:recap} Notation used to describe filtering.  This table
provides a quick reference for symbols used.}
\end{table}

In table \ref{table:flops} we present the computational cost scaling in
floating point operations per sample for common tasks in the pipeline.  
%
% common tasks FLOPS table
%
\begin{table}[htdp]
\begin{center}
\begin{tabular}{l l}
\bf{Process} & \bf{ops/sample} \\
\hline
% FIR
\fir\ matched filter, \numtmps\ templates of length \tmpsamps\ & $2 \numtmps \tmpsamps$ \\
% FFT
\fft\ matched filter, \numtmps\ templates of length \tmpsamps\, & \multirow{2}{*}{$\mbox{\fontsize{14}{17}\selectfont $\frac{4 (\numtmps + 1) \lg \fftblock + 2 \numtmps}{1 - \tmpsamps/\fftblock}$}$} \\
% FFT contd...
$\,\,\,\,\,\,$ blocks of length \fftblock & \\
% Resampling
\fir\ resampling filter, length \resampsamps\ for each of \numtmps\ templates& \multirow{2}{*}{$2 \numtmps \resampsamps\ f_1 / f_2$} \\
% Resampling contd...
$\,\,\,\,\,\,$ and sample rates $f_1 < f_2$ & \\
% Matrix multiply
multiply $M \times L$ real matrix by $L\times1$ real vector & $2 M L$ \\
%
\end{tabular}
\end{center}
\caption{Number of floating point operations per sample (multiplications and
divisions) required for a selection of signal processing operations used in
\textsc{lloid}.}
\label{table:flops}
\end{table}

The filter bank can be implemented using finite impulse response \fir\ filters,
which are just sliding window dot products.  If there are \numtmps\ templates
of length \tmpsamps\, and the data stream contains \hoftsamps\ samples, then
applying the filter bank requires $2 \numtmps \tmpsamps \hoftsamps$ operations.

More commonly, the matched filters are implemented using the \fft\ convolution.
This entails applying {\fft}s to blocks of \fftblock\ samples, with $\tmpsamps
\leq \fftblock$, each block overlapping the previous one by $ \fftblock -
\tmpsamps $ samples.  There are $\hoftsamps/(\fftblock-\tmpsamps)$ such blocks
required to filter \hoftsamps\ samples of data.  Modern implementations of the
Cooley-Tukey \fft, such as the ubiquitous \texttt{fftw}, require about $4 N \lg
N$ operations to evaluate a \textsc{dft} of size $N$~\cite{Johnson:2007p9654}.
%
%
\editorial{This is more commonly known as ``overlap-save''.  We should find
someone else's operation count and cite it.}  
%
%
A \fftblock\ sample cross-correlation consists of a forward \fft, a \fftblock
sample dot product, and an inverse \fft\, totaling $8 \fftblock \lg \fftblock +
2 D$ operations per block.  Per sample, this is $(8 \lg \fftblock + 2) / (1 -
\tmpsamps/\fftblock)$ operations.  As this expression indicates the number
of operations increases as the block size \fftblock\ approaches the
filter length \tmpsamps.
%
%
\editorial{Drew: Why don't we change this to an overlap of $m$ samples so we
can see what happens as we increase the overlap to reduce latency.}

The \fir\ filter implementation has the advantage that it has no intrinsic
latency, whereas the \fft\ convolution has latency of $\fftblock-\tmpsamps$. 
%
%
%For example, for a $1.4 -
%1.4 \, M_\odot$ template with duration $\sim 1 \, \mathrm{ks}$, the
%\textsc{fft} convolution has a latency $\geq 2 \, \mathrm{ks}$.  
%
However, the \fir\ filter implementation has the disadvantage of much greater
overhead per sample than the \fft\ convolution.  For a $1\,\mathrm{ks}$
template sampled at $4096\,\mathrm{Hz}$, the \textsc{fir} implementation
requires about about $\tmpsamps / 8 \lg 2 \tmpsamps = 2.2 \times 10^4$ times
more operations per sample than the \fft implementation.  We will now consider
the computational cost of the \fir\ filter implementation described in the
previous sections.

It is convenient to express the computational cost of the entire filtering
procedure in floating point operations per second {\flops}.  The cost will be
the sum of the cost of the \fir\ filtering for the orthogonal filter in each
time slice plus the cost of reconstructing the original waveforms with matrix
operations and resampling.  Using the formulas in table \ref{table:flops} we
arrive at
%
\begin{equation}
%\mathrm{Computational\, cost} 
\mathrm{C.C.} \propto 2 \sum_k^{\numslices} (\slicessamps + \numtmps) \numsvdtmps f_k \nonumber
	r \sum_{k, f_k \neq f_{\fmax}}^{\numslices} \numtmps \resampsamps f_k 
\label{eq:TDcost}
\end{equation}
%
where the first sum is the total cost for filtering and reconstructing the
orthogonal filters $u_{jk}$.  The second sum is the cost of resampling the
reconstructed time slice outputs to the original sample rate {\fmax}.  It
assumes a \fir\ filter with \resampsamps\ sample points.  Note that the cost
for resampling only occurs for the downsampled time slices. The resampling cost
largely depends on {\resampsamps}.  The computational cost of \eqref{eq:TDcost}
is dominated by the highest frequency terms in the sum.  Comparing just the
highest frequency term of \eqref{eq:TDcost} with \eqref{eq:SNRTD} shows \order{
(\slicesamps{\mathrm{max}} + \numtmps) \svdtmps{\mathrm{max}} \fmax} \flops\
versus \order{\numtmps \tmpsamps \fmax} {\flops}.  The full calculation for a
particular patch of parameter space is shown in table \ref{table:comp_cost}.
\hl{FIXME we need to actually say what went into this}

\editorial{From here down introduce an example, but don't go through the FFT
methods in lloid focus just on the comparison of the TD method without lloid
tricks, the FD method without lloid tricks, and the TD lloid method (no comp
detection statistic though), remember that the goal is near realtime detection
not computational cost improvements on FD methods}

\editorial{It would be good to illustrate the layout of this particular
template bank: masses spanned, time slice layout \dots}
\begin{table}[!h]
\label{table:comp_cost}
\begin{center}
\input{flop_budget_example}
\end{center}
\caption{\hl{FIXME: Operation counts per sample for four different detection methods.}}
\end{table}
